Rapid slip-deficit rates at the eastern margin of the Tibetan plateau prior to the 2008 $\textrm{M}_\textrm{w}$ 7.9 Wenchuan earthquake

### 2014-11-14 paper.tex
Incorporated JS edits.

\title{Rapid slip-deficit rates at the eastern margin of the Tibetan plateau prior to the 2008 $\textrm{M}_\textrm{w}$ 7.9 Wenchuan earthquake}
\author{T. Ben Thompson, Brendan J. Meade}
\date{September 2014}
\usepackage{graphicx}
\usepackage[margin=1.0in]{geometry}
\usepackage[nomarkers,figuresonly]{endfloat}
\usepackage{setspace}
\usepackage{natbib}
\doublespacing

@@

\section{Introduction}
The Longmen Shan range is the steepest margin of the Tibetan Plateau rising 5 km over a distance of only 30 km at the western edge of the Sichuan Basin. In 2008, the $\textrm{M}_{\textrm{w}}$ 7.9 Wenchuan earthquake ruptured the Beichan and Penguan rangefront thrusts ${\sim}250$ km along strike, leading to approximately 70,000 fatalities. The energy release associated with the Wenchuan earthquake requires a complimentary period of strain accumulation. However, interseismic GPS velocities measured prior to the occurence of the Wenchuan earthquake have been interpreted as indicating neglible shortening \citep{king97, chen00, shen05, Meade07c, Loveless2011}. This apparent lack of present-day shortening across the steepest topography of the India-Asia collision has been hypothesized to result from lower crustal inflation tectonic model for the eastern Tibetan Plateau \citep{royden97, bird91, Burchfiel2008, Clark2000}.
But, these models do not provide a mechanism for accumulating the elastic strain released coseismically during the Wenchuan earthquake. Recent observations of co- and post-seismic slip on a deep detachment to the west of the Beichuan fault\citep{Qi2011, Fielding2013b} motivate a new model of Longmen Shan interseismic shortening that includes a locked detachment.

Previous estimates of geodetically constrained shortening and slip-deficit rates in the Longmen Shan region can be categorized into those that either: 1) do not explicitly treat earthquake cycle effects, \citep{chen00, shen05, Thatcher2007} or 2) those that use block models, with a first order quasi-static approximation of earthquake cycle effects but coarse representations of fault system geometry \citep{Meade07c, Loveless2011, Burchfiel2008}. The inferred slip-deficit rates range from 0.0 mm/yr \citep{Thatcher2007} to 3.2 mm/yr \citep{Loveless2011}, with all analyses reporting significant strain 100-200 km northwest of the Longmen Shan in a region termed the Songpan-Xihe deformation zone \citep{shen05}. Here, we present results from a two-dimensional boundary element model that includes earthquake cycle and topographic effects to explain the regional geodetic velocities as the result of interseismic locking on both the range-front Beichuan fault and a 20 km deep detachment beneath the hinterland. This fault system geometry is consistent with the inference of 2-6 m of slip of a 20 km deep detachment that extends 90 km northwest using InSAR measurements and post-Wenchuan GPS observations \citep{Qi2011, Fielding2013b} and interpretation of structural observations \citep{Hubbard2010, Li2010a}. Integrating rangefront and detachment fault system geometry, we develop a model of interseismic strain accumulation that is consistent with both interseismic GPS velocities measured prior to the Wenchuan earthquake and previous observations of co- and post-seismic strain release at depth. Importantly, the model provides a mechanism for large slip-deficit rates in the abscence of a localized velocity gradient at the Longmen Shan rangefront.

@@

\section{Geodetic analysis of Longmen Shan shortening and slip deficit rates}
The geometry of the Beichaun fault and the detachment to the west have been inferred from 1) structural interpretations  \citep{Hubbard2010} (Plesch and Shaw, personal communication), 2) seismic relection observations \citep{Zhang2009, Guo2013}, and 3) geodetic modeling of co- and post-seismic deformation associated with the Wenchuan earthquake \citep{Qi2011, Fielding2013b}. The depth and dip of the detachment are similar to the best-fitting surface found by \citet{Qi2011}. The fault system geometry consists of two main segments (Figure \ref{fig:regional_map}). The range-front Beichuan fault dips steeply at ${\sim}60^{\circ}$ at its surface trace shallowing to sub-horizontal ${\sim}20$ km depth where it merges with the detachment surface. The detachment dips at ${\sim}1.5^{\circ}$ from an initial depth of 20 km and extends 110 km to the northwest to a depth of 23 km. For our interseismic shortening and slip-deficit models we use an idealized two-dimensional representation of these fault surfaces (Figure \ref{fig:regional_map}).

@@

The velocity gradient across the Longmen Shan rangefront is negligible (Figure \ref{fig:big_stack}a). To fit these observed near-field data, a Beichuan-only fault geometry, dipping ${\sim}60^{\circ}$ at the surface and shallowing to ${\sim}10^{\circ}$ at a maximum locking depth of 20 km would require near-zero slip-deficit rates. Such a model is inconsistent with slip during the Wenchuan earthquake on the Beichuan and Penguan faults. It is also inconsistent with co- and post-seismic slip on the detachment \citep{Qi2011, Fielding2013b}.

@@

The primary uncertainty in our shortening estimate is the lack of good constraints on the detachment geometry in the hinterland. However, assuming the geometry is correct, two further forms of error are evident. First, the uncertainty in true interseismic velocities is propagated through to the estimated shortening, resulting in the gray region of uncertainty in Figure \ref{fig:big_stack}a. Further, the sparsity of observations suggests that certain observations may have undue influence on the shortening estimate. Excluding the northwest-most observation decreases the best-fit shortening to 3.9 $\pm$ 1.9 mm/yr, with a near-surface slip-deficit rate of 6.5 (give uncertainty) mm/yr on the Beichuan fault. The least squares uncertainty is increased (from 1.5 mm/yr to 1.9 mm/yr) because the northwestern portion of velocities are poorly constrained once the observation is excluded. This northwestern-most observation is adjacent to the Longriba fault, suggesting that it should not be included. However, the Longriba fault is primarily a dextral strike-slip fault \citep{Ren2013}, so it's influence should be predominantly orthogonal to the profile parallel velocities we examine here.

@@

\section{Implications for the Longmen Shan earthquake cycle}
The fundamental effect of allowing a 20 km detachment to accumulate and release elastic strain through the earthquake cycle is the lack of a range-front geodetic signature despite the significant ($5-10$ mm/yr) slip-deficit rate (\ref{fig:big_stack}a). Locking on the deep detachment obscures the signal of deformation on the range-front faults. The interseismic deformation is much more broadly distributed across more than 300 kilometers. According to this model, interseismically, the eastern Tibetan Plateau moves cohesively with the Sichuan Basin, whereas coseismic slip is concentrated at the boundary between these two blocks.

@@

While we have explained the observed the broad GPS interseismic velocity gradient as a result of a locked fault geometry including a deep detachment, we have not addressed when accumulated slip-deficit on the deep detachment is released. Geodetically constrained estimates of co- and post-seismic slip shows up to 5 m of slip on a deep detachment \citep{Qi2011}. Our data analysis depends on the detachment being locked during the period of data collection from 1998 to 2004 \citep{gan07}. In addition to coseismic and short-term postseismic activity on the detachment, it is possible that intermittent aseismic slip could be an important mode of strain release. This is difficult to observationally constrain given that intersimsic GPS observations cover, at best, 6\% of a complete Longmen Shan earthquake cycle. Seismicity is approximately limited to the upper 20 km of crust in the range-front \citep{Li2010a}. Using steady-state critical taper wedge theory \citep{dahlen90}, with a internal coefficient of friction of 0.7, a cohesion of 5 MPa, and a hydrostatic fluid pressure, we find a basal friction coefficient of 0.08. This is comparable to estimates of friction on the low-angle detachment system beneath the Longmen Shan foreland \citep{Hubbard2010}, indicating a very weak fault over long time-periods. On the other hand, the surface area and shortening rate of the detachment could sustain very large earthquakes if slip is released coseismically. Specifically, if 100 km down-dip and 300 km along-strike were to slip 4 meters, similar to average slip during the Wenchuan earthquake, a $\textrm{M}_{\textrm{w}}$ 8.4 rupture would be produced.

The 6-9 mm/yr slip-deficit rates estimated here can be used to provide a new kinematic constraint on earthquake recurrence intervals at the Longmen Shan (Figure \ref{fig:hazard}). Our shortening estimate suggests a recurrence interval for Wenchuan-like ruptures of approximately 600 years, with much shorter recurrence intervals for smaller events. This contrasts with the average recurrence of ~2000 years inferred from paleoseismic field investigations for the Beichuan fault \citep{Ran2010}. The 6-9 mm/yr slip-deficit rate estimated here would imply $\textrm{M}_{\textrm{w}}$ 9.0 ruptures for that recurrence interval, which would seemingly require an along strike rupture equal to or greater than the ${\sim}300$ km along strike length of the Longmen Shan. However, our slip-deficit estimate could include the effects of other imbricated structures. As a result our estimate is a system-wide estimate while paleoseismic estimates have focused on the analysis of individual thrust faults.

>>>>

\title{Rapid slip-deficit rates at the eastern margin of the Tibetan plateau prior to the 2008 $\textrm{M}_\textrm{w}$ 7.9 Wenchuan earthquake}
\author{T. Ben Thompson, Brendan J. Meade -- where to put JS, AP?}
\usepackage{graphicx}
\usepackage[margin=1.0in]{geometry}
\usepackage[nomarkers,figuresonly]{endfloat}
\usepackage{setspace}
\usepackage{natbib}
\doublespacing

@@

\section{Introduction}
The Longmen Shan range is the steepest margin of the Tibetan Plateau rising 5 km over a distance of only 30 km at the western edge of the Sichuan Basin. In 2008, the $\textrm{M}_{\textrm{w}}$ 7.9 Wenchuan earthquake ruptured the Beichan and Penguan rangefront thrusts ${\sim}250$ km along strike, leading to approximately 70,000 fatalities \citep{Hubbard2009, Xu2009, Lin2009}. The energy release associated with the Wenchuan earthquake requires a complimentary period of strain accumulation. However, interseismic GPS velocities measured prior to the occurence of the Wenchuan earthquake have been interpreted as indicating neglible shortening \citep{king97, chen00, shen05, Meade07c, Loveless2011}. This apparent lack of present-day shortening across the steepest topography of the India-Asia collision has been hypothesized to result from lower crustal inflation tectonic model for the eastern Tibetan Plateau \citep{royden97, bird91, Burchfiel2008, Clark2000}. But, these models do not provide a mechanism for accumulating the elastic strain released coseismically during the Wenchuan earthquake, nor do they explain the significant crustal shortening and active geologic structures along the range front \citep{Hubbard2010, Li2010a, Wang2013, Wang2014}. Recent observations of co- and post-seismic slip on a deep detachment to the west of the Beichuan fault\citep{Qi2011, Fielding2013b} motivate a new model of Longmen Shan interseismic shortening that includes a locked detachment.

Previous estimates of geodetically constrained shortening and slip-deficit rates in the Longmen Shan region can be categorized into those that either: 1) do not explicitly treat earthquake cycle effects, \citep{chen00, shen05, Thatcher2007} or 2) those that use block models, with a first order quasi-static approximation of earthquake cycle effects but coarse representations of fault system geometry \citep{Meade07c, Loveless2011, Burchfiel2008}. The inferred slip-deficit rates range from 0.0 mm/yr \citep{Thatcher2007} to 3.2 mm/yr \citep{Loveless2011}, with all analyses reporting significant strain 100-200 km northwest of the Longmen Shan in a region termed the Songpan-Xihe deformation zone \citep{shen05}. Here, we present results from a two-dimensional boundary element model that includes earthquake cycle and topographic effects to explain the regional geodetic velocities as the result of interseismic locking on both the range-front Beichuan fault and a 20 km deep detachment beneath the hinterland. This fault system geometry is consistent with structural observations \citep{Hubbard2010, Li2010a} and the inference of 2-6 m of slip of a 20 km deep detachment that extends 90 km northwest using InSAR measurements and post-Wenchuan GPS observations \citep{Qi2011, Fielding2013b}. Integrating rangefront and detachment fault system geometry, we develop a model of interseismic strain accumulation that is consistent with both interseismic GPS velocities measured prior to the Wenchuan earthquake and previous observations of co- and post-seismic strain release at depth. Importantly, the model provides a mechanism for large slip-deficit rates in the abscence of a localized velocity gradient at the Longmen Shan rangefront.

@@

\section{Geodetic analysis of Longmen Shan shortening and slip deficit rates}
The geometry of the Beichuan fault and the detachment to the west has been inferred from structural interpretations \citep{Hubbard2010} and seismic reflections observations \citep{Zhang2009, Guo2013}. The Beichuan fault is part of an imbricate thrust system, with fault dips steepening to the northwest. Structural imbrication is typically produced by stacked listric faults that sole to detachments (TODO: CITE?). Furthermore, the large synclinorium parallel to, and west of, the Beichuan fault is consistent with the Beichuan fault shallowing to a detachment based on fault-bend folding theory \citep{Hubbard2010, Suppe1983}. In our model, the range-front Beichuan fault dips steeply at ${\sim}60^{\circ}$ at its surface trace shallowing to sub-horizontal ${\sim}20$ km depth where it merges with the detachment surface. At seismogenic depths, a 35$^\circ$ dip has been confirmed via focal mechanisms \citep{Hubbard2010}. Extrapolating from the regional dip of the Sichuan Basin as it extends beneath the Longmenshan range front \citep{Hubbard2010, Guo2013}, the detachment in our model dips ${\sim}1.5^{\circ}$ from an initial depth of 20km to a depth of 23km at a distance of 110km to the northwest. The depth, dip, and extent of the detachment are similar to the best-fitting surface found by \citet{Qi2011} in fitting the co- and post-seismic slip of the Wenchuan earthquake. For our interseismic shortening and slip-deficit models we use an idealized two-dimensional representation of these fault surfaces (Figure \ref{fig:regional_map}).

@@

The velocity gradient across the Longmen Shan rangefront is negligible (Figure \ref{fig:big_stack}a). To fit these observed, any planar or listric geometry with interseismic locking limited to the range-front (for example, a Beichuan-only fault geometry) would require near-zero slip-deficit rates. Such a model is inconsistent with slip during the Wenchuan earthquake on the Beichuan and Penguan faults and significant geologic estmates of shortening \citep{Hubbard2010, Li2010, Wang2013, Wang2014}, and . It is also inconsistent with co- and post-seismic slip on the detachment \citep{Qi2011, Fielding2013b}.

@@

The primary uncertainty in our shortening estimate is the lack of good constraints on the detachment geometry in the hinterland. However, assuming the geometry is correct, two further forms of error are evident. First, the uncertainty in true interseismic velocities is propagated through to the estimated shortening, resulting in the gray region of uncertainty in Figure \ref{fig:big_stack}a. Further, the sparsity of observations suggests that certain observations may have undue influence on the shortening estimate. Excluding the northwest-most observation decreases the best-fit shortening to 3.9 $\pm$ 1.9 mm/yr, with a near-surface slip-deficit rate of 6.5 (give uncertainty) mm/yr on the Beichuan fault. The least squares uncertainty is increased (from 1.5 mm/yr to 1.9 mm/yr) because the northwestern portion of velocities are poorly constrained once the observation is excluded. This northwestern-most observation is adjacent to the Longriba fault, and thus may be impacted by interseismic strain accumulation on this structure. However, the Longriba fault is primarily a dextral strike-slip fault \citep{Ren2013}, so it's influence should be predominantly orthogonal to the profile parallel velocities we examine here.

@@

\section{Implications for the Longmen Shan earthquake cycle}
The fundamental effect of allowing a 20 km deep detachment to accumulate and release elastic strain through the earthquake cycle is the lack of a range-front geodetic signature despite the significant ($5-10$ mm/yr) slip-deficit rate (\ref{fig:big_stack}a). Locking on the deep detachment obscures the signal of deformation on range-front faults. The interseismic deformation is much more broadly distributed across over 300 kilometers. According to this model, interseismically, the easternmost Tibetan Plateau moves cohesively with the Sichuan Basin, whereas coseismic slip is concentrated at the boundary between these two blocks.

@@

While we have explained the observed broad GPS interseismic velocity gradient as a result of a locked fault geometry including a deep detachment, we have not addressed when accumulated slip-deficit on the deep detachment is released. Geodetically constrained estimates of co- and post-seismic slip shows up to 5 m of slip on a deep detachment \citep{Qi2011}. Our data analysis depends on the detachment being locked during the period of data collection from 1998 to 2004 \citep{gan07}. In addition to coseismic and short-term postseismic activity on the detachment, it is possible that intermittent aseismic slip could be an important mode of strain release. This is difficult to observationally constrain given that intersimsic GPS observations cover, at best, 6\% of a complete Longmen Shan earthquake cycle. Seismicity is approximately limited to the upper 20 km of crust in the range-front \citep{Li2010a}. Using steady-state critical taper wedge theory \citep{dahlen90}, with a internal coefficient of friction of 0.7, a cohesion of 5 MPa, and a hydrostatic fluid pressure, we find a basal friction coefficient of 0.08. This is comparable to estimates of friction on the low-angle detachment system beneath the Longmen Shan foreland \citep{Hubbard2010}, indicating a very weak fault over long time-periods. On the other hand, the surface area and shortening rate of the detachment could sustain very large earthquakes if slip is released coseismically. Specifically, if 100 km down-dip and 300 km along-strike were to slip 4 meters, similar to average slip during the Wenchuan earthquake, a $\textrm{M}_{\textrm{w}}$ 8.4 rupture would be produced.

The 6-9 mm/yr slip-deficit rates estimated here can be used to provide a new kinematic constraint on earthquake recurrence intervals at the Longmen Shan (Figure \ref{fig:hazard}). Our shortening estimate suggests a recurrence interval for Wenchuan-like ruptures between 470 and 800 years (best fit: 590 years), with much shorter recurrence intervals for smaller events. This contrasts with the average recurrence of ~2000 years inferred from paleoseismic field investigations for the Beichuan fault \citep{Ran2010}. The 6-9 mm/yr slip-deficit rate estimated here would imply $\textrm{M}_{\textrm{w}}$ 9.0 ruptures for that recurrence interval, which would seemingly require an along strike rupture equal to or greater than the ${\sim}300$ km along strike length of the Longmen Shan. However, our slip-deficit estimate could include the effects of other imbricated structures, such as the Range Front blind thrust \citep{Hubbard2010, Wang2014}. As a result our estimate is a system-wide estimate while paleoseismic estimates have focused on the analysis of individual thrust faults.
